---
title: "Mizer-Fish-Model"
author: "Bethan Sleath"
date: "2026-02-21"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(mizer)
library(mizerExperimental)
library(tidyverse)
library(plotly)
library(dplyr)
```

## Setting up model

We set the parameters of the model to be a single species model with semi-chemostatic resource dynamics. Initial conditions for resource and fish abundance are altered.

```{r}
params_single <- newSingleSpeciesParams()
params_initial <- setResource(params_single,
                      resource_dynamics = "resource_semichemostat")
params_double <- params_initial
params_double <- setResource(params_double,
                             resource_level = 0.05)

initialN(params_double) <- 2*initialN(params_double)
initialNResource(params_double) <- initialNResource(params_double)/2
```

## Fishing gear

```{r}
gear_params(params_double) <- data.frame(
  gear = "gear",
  species = "Target species",
  catchability = 0.8,
  sel_func = "sigmoid_weight",
  sigmoidal_weight = 15,
  sigmoidal_sigma = 5)

params_double <- setFishing(params_double, gear_params = gear_params)

gear_params(params_double)
```

## Locally reduce the resource rate and resource capacity

### Resource rate

```{r}
rr <- resource_rate(params_double)
w_full <- w_full(params_double)
rr[192:232] <- rr[192:232] / 2

rr_data <- data.frame(
  Weight = params_double@w_full,
  rr = rr)

rr_plot <- ggplot(rr_data, 
                  aes(x = Weight, y = rr)) +
  geom_smooth() +
  scale_x_log10() + 
  scale_y_log10() +
  theme_classic()
rr_plot
```

### Resource capacity

```{r}
rc <- resource_capacity(params_double)
rc[192:232] <- rc[192:232] / 5

rc_data <- data.frame(
  Weight = params_double@w_full,
  rc = rc
)

rc_plot <- ggplot(rc_data, 
                  aes(x = Weight, y = rc)) +
  geom_smooth() +
  scale_x_log10() + 
  scale_y_log10() +
  theme_classic()
rc_plot
```

### Set new parameters with reductions local reductions

```{r}
params_reduced <- setResource(params_double,
                              resource_capacity = rc,
                              resource_rate = rr,
                              balance = FALSE)
```

## Narrow predation kernel

```{r}
pred_kernel <- getPredKernel(params_reduced)
pred_kernel_reduced <- pred_kernel[, 89, , drop = FALSE]

ggplot(melt(pred_kernel_reduced)) +
  geom_line(aes(x = w_prey, y = value)) +
  scale_x_log10(limits = c(1e-4, 100))

select(species_params(params_reduced), beta, sigma)

params <- params_reduced

given_species_params(params)$sigma <- 0.8
given_species_params(params)$beta <- 1000

getPredKernel(params)[, 89, , drop = FALSE] %>% 
  melt() %>% 
  ggplot() +
  geom_line(aes(x = w_prey, y = value)) +
  scale_x_log10(limits = c(1e-4, 100))
```

## Calculate flux of system

```{r}
sim_reduced <- project(params, t_max = 30, effort = 1)

N_reduced <- finalN(sim_reduced)["Target species", , drop = TRUE]
w <- w(params_reduced)

E_growth_reduced <- getEGrowth(params)["Target species", , drop = TRUE]
grr <- w * E_growth_reduced
flux_reduced <- grr * N_reduced


reduced_flux_data <- data.frame(Weight = w, 
                             Flux = flux_reduced)
reduced_flux_data$species = "Target Species"

plot_ly(reduced_flux_data) |> 
  add_lines(x = ~Weight, y = ~Flux, color = ~species) |> 
  layout(yaxis = list(type = "log", exponentformat = "power",
                              title_text = "Flux (g/year)"),
         xaxis = list(type = "log", exponentformat = "power", 
                      title_text = "Weight (g)"))
```

## Plot flux over limited weight range

```{r}
limited_flux_data <- data.frame(Weight = w[89:101], 
                                Flux = flux_reduced[89:101])
limited_flux_data$Species = "Target Species"

plot_ly(limited_flux_data) |> 
  add_lines(x = ~Weight, y = ~Flux, color = ~Species) |> 
  layout(yaxis = list(type = "log", exponentformat = "power",
                      title_text = "Flux (g/year)"),
         xaxis = list(type = "log", exponentformat = "power", 
                      title_text = "Weight (g)"))
```

## Plot growth rate

```{r}
growth_data <- data.frame(growth = E_growth_reduced,
                          weight = w)
growth_data$Species = "Target Species"

plot_ly(growth_data) |> 
  add_lines(x = ~weight, y = ~growth, color = ~Species) |> 
  layout(yaxis = list(#type = "log", exponentformat = "power",
                      title_text = "Growth Rate (g/year)"),
         xaxis = list(#type = "log", exponentformat = "power", 
                      title_text = "Weight (g)"))
```

## Plot growth rate over limited weight range
```{r}
limited_growth_data <- data.frame(growth = E_growth_reduced[89:101],
                          weight = w[89:101])
limited_growth_data$species = "Target Species"
plot_ly(limited_growth_data) |> 
  add_lines(x = ~weight, y = ~growth, color = ~species) |> 
  layout(yaxis = list(#type = "log", exponentformat = "power",
    title_text = "Growth Rate (g/year)"),
    xaxis = list(#type = "log", exponentformat = "power", 
      title_text = "Weight (g)"))
```

## biomass spectra
```{r}
plotSpectra(sim_reduced, power = 2, wlim = c(1e-8, NA), ylim = c(1e-8, NA),
            time_range = 30)
```

## Animation of biomass abundance
```{r}
nf <- melt(sim_reduced@n)
n_ppf <- melt(sim_reduced@n_pp)
n_ppf$sp <- "Resource"
nf <- rbind(nf, n_ppf)

plot_ly(nf) %>%
  # show only part of plankton spectrum
  filter(w > 10^-5) %>% 
  # start at time 20
  #filter(time >= 26) %>% 
  # calculate biomass density with respect to log size
  mutate(b = value * w^2) %>% 
  # Plot lines
  add_lines(
    x = ~w, y = ~b,
    color = ~sp,
    frame = ~time,
    line = list(simplify = FALSE)
  ) %>% 
  # Use logarithmic axes
  layout(params, xaxis = list(type = "log", exponentformat = "power",
                         title_text = "body mass (g)"),
         yaxis = list(type = "log", exponentformat = "power",
                      title_text = "biomass (g/m^3)",
                      range = c(-8, 0)))
```

## Animation of flux
```{r}
species <- "Target species"
time <- getTimes(sim_reduced)

flux_series <- lapply(seq_along(time), function(i) {
  n_at_t <- sim_reduced@n[i, species, ]
  data.frame(
    Time = time[i],
    Weight = w,
    Flux = n_at_t * grr)
}) %>% bind_rows()

flux_series$Species <- species

plot_ly(flux_series) %>% 
  add_lines(x ~Weight, 
            y ~Flux, 
            color = ~Species, 
            frame = ~Time, 
            line = list(simplify = FALSE)) %>%  
  layout(params, xaxis = list(type = "log", exponentformat = "power",
                         title_text = "body mass (g)"),
         yaxis = list(type = "log", exponentformat = "power",
                      title_text = "flux (g/year)", 
                      range = c(-8, 2)))
```

