---
title: "Mizer-Fish-Model"
author: "Bethan Sleath"
date: "2026-02-21"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(mizer)
library(mizerExperimental)
library(tidyverse)
library(ggplot2)
library(plotly)
library(dplyr)
```

## Setting up model

We set the parameters of the model to be a single species model with semi-chemostat growth resource dynamics. Initial conditions for resource and fish abundance are altered.

```{r message=FALSE, warning=FALSE}

params_single <- newSingleSpeciesParams() 
params_single <- setResource(params_single, 
                             resource_dynamics = "resource_semichemostat")

params_initial <- setResource(params_single,
                              resource_level = 0.05)

initialN(params_initial) <- 2*initialN(params_initial)
initialNResource(params_initial) <- initialNResource(params_initial)/2
```

## Fishing gear

Add a fishing gear to the model which uses a sigmoidal weight selectivity function

```{r message=FALSE, warning=FALSE}
gear_params(params_initial) <- data.frame(
  gear = "gear",
  species = "Target species",
  catchability = 1,
  sel_func = "sigmoid_weight",
  sigmoidal_weight = 25,
  sigmoidal_sigma = 10)

params_initial <- setFishing(params_initial, gear_params = gear_params)

gear_params(params_initial)
```

## Initial model

### Maximum sustainable yield
We calculate the maximum sustainable yield of the initial system in order to determine the fishing effort of the simulation

```{r message=FALSE, warning=FALSE}
plotYieldVsF(
  params_single,
  'Target species',
  F_range,
  F_max = 1,
  F_min = 0,
  no_steps = 25,
  distance_func = distanceSSLogN,
  tol = 0.001,
  t_max = 250) +
  theme_test()
```

### Calculate flux
Calculate the flux, the movement biomass between size classes, of the original system

```{r message=FALSE, warning=FALSE}
sim_initial <- project(params_initial, t_max = 50, effort = 0.125)

N <- finalN(sim_initial)["Target species", , drop = TRUE]
w <- w(params_initial)

E_growth <- getEGrowth(params_initial)["Target species", , drop = TRUE]
gr <- w * E_growth
flux <- gr * N

initial_flux_data <- data.frame(Weight = w, 
                                Flux = flux)

plot_ly(initial_flux_data) |> 
  add_lines(x = ~Weight, y = ~Flux) |> 
  layout(yaxis = list(type = "log", exponentformat = "power",
                      title_text = "Flux (gm<sup>-2</sup>year<sup>-1</sup>)", 
                      showline = TRUE, showgrid = FALSE, mirror = TRUE),
         xaxis = list(type = "log", exponentformat = "power", 
                      title_text = "Weight (g)", 
                      showline = TRUE, showgrid = FALSE, mirror = TRUE), 
         margin = list(b = 65, l = 50))
```

### Flux over a limited weight range

```{r message=FALSE, warning=FALSE}
initial_limited_flux <- data.frame(Weight = w[61:101], 
                                   Flux = flux[61:101])

plot_ly(initial_limited_flux) |> 
  add_lines(x = ~Weight, y = ~Flux) |> 
  layout(yaxis = list(type = "log", exponentformat = "power",
                      title_text = "Flux (gm<sup>-2</sup>year<sup>-1</sup>)",
                      showline = TRUE, showgrid = FALSE, mirror = TRUE),
         xaxis = list(type = "log", exponentformat = "power", 
                      title_text = "Weight (g)",
                      showline = TRUE, showgrid = FALSE, mirror = TRUE),
         margin = list(b = 65, l = 50))
```

## System with reduced resource availability

### Locally reduce the resource rate and resource capacity
In order to model high competition between predators and a high density of fish within size classes, we locally reduce the resource rate and capacity to limit resource availability.

### Resource rate

```{r message=FALSE, warning=FALSE}
rr <- resource_rate(params_initial)
w_full <- w_full(params_initial)
rr[182:242] <- rr[182:242] / 2

rr_data <- data.frame(
  Weight = params_initial@w_full,
  rr = rr)

rr_plot <- ggplot(rr_data, 
                  aes(x = Weight, y = rr)) +
  geom_line() +
  scale_x_log10() + 
  scale_y_log10() +
  theme_classic()
rr_plot
```

### Resource capacity

```{r message=FALSE, warning=FALSE}
rc <- resource_capacity(params_initial)
rc[182:242] <- rc[182:242] / 5

rc_data <- data.frame(Weight = params_initial@w_full,
                      rc = rc)

rc_plot <- ggplot(rc_data, 
                  aes(x = Weight, y = rc)) +
  geom_line() +
  scale_x_log10() + 
  scale_y_log10() +
  theme_classic()
rc_plot
```

### Set new parameters with reductions local reductions

```{r message=FALSE, warning=FALSE}
params_reduced <- setResource(params_initial,
                              resource_capacity = rc,
                              resource_rate = rr,
                              balance = FALSE)
```

## Narrow predation kernel
Another parameter we alter to model high competition is narrowing the predation kernel. This increases the specificity of predators food preference, and limits resource abundance.

```{r message=FALSE, warning=FALSE}
pred_kernel <- getPredKernel(params_reduced)

pred_kernel_reduced <- pred_kernel[, 89, , drop = FALSE]

ggplot(melt(pred_kernel_reduced)) +
                            geom_line(aes(x = w_prey, y = value)) +
                            scale_x_log10("Weight of prey (g)", limits = c(1e-5, 1e03)) +
                            scale_y_continuous("Proportion", limits = c(0,1)) +
                            theme_bw()


select(species_params(params_reduced), beta, sigma)

params <- params_reduced

given_species_params(params)$sigma <- 0.8
given_species_params(params)$beta <- 1000

getPredKernel(params)[, 89, , drop = FALSE] %>% 
  melt() %>% 
  ggplot() +
  geom_line(aes(x = w_prey, y = value)) +
  scale_x_log10("Weight of prey (g)", limits = c(1e-5, 1e03)) +
  scale_y_continuous("Proportion")+
  theme_bw()
```

### Maximum sustainable yield
We once again calculate the maximum sustainable yield of the system to determine the fishing effort of the simulation

### Calculate flux of system

```{r message=FALSE, warning=FALSE}
sim_reduced <- project(params, t_max = 50, effort = 0.75)

N_reduced <- finalN(sim_reduced)["Target species", , drop = TRUE]
w <- w(params)

E_growth_reduced <- getEGrowth(params)["Target species", , drop = TRUE]
grr <- w * E_growth_reduced
flux_reduced <- grr * N_reduced


reduced_flux_data <- data.frame(Weight = w, 
                             Flux = flux_reduced)

plot_ly(reduced_flux_data) |> 
  add_lines(x = ~Weight, y = ~Flux) |> 
  layout(yaxis = list(type = "log", exponentformat = "power",
                      title_text = "Flux (gm<sup>-2</sup>year<sup>-1</sup>)", 
                      showline = TRUE, showgrid = FALSE, mirror = TRUE),
         xaxis = list(type = "log", exponentformat = "power", 
                      title_text = "Weight (g)", 
                      showline = TRUE, showgrid = FALSE, mirror = TRUE),
         margin = list(b = 65, l = 50))
```

### Flux over a limited weight range

```{r message=FALSE, warning=FALSE}
limited_flux_data <- data.frame(Weight = w[61:101], 
                                Flux = flux_reduced[61:101])

plot_ly(limited_flux_data) |> 
  add_lines(x = ~Weight, y = ~Flux) |> 
  layout(yaxis = list(type = "log", exponentformat = "power",
                      title_text = "Flux (gm<sup>-2</sup>year<sup>-1</sup>)",
                      showline = TRUE, showgrid = FALSE, mirror = TRUE),
         xaxis = list(type = "log", exponentformat = "power", 
                      title_text = "Weight (g)",
                      showline = TRUE, showgrid = FALSE, mirror = TRUE),
         margin = list(b = 65, l = 50))
```

## Growth rate

### Reduced resource system
We plot the growth rate of the fish population in the ecosystem.

```{r message=FALSE, warning=FALSE}
growth_data <- data.frame(growth = E_growth_reduced,
                          weight = w)

plot_ly(growth_data) |> 
  add_lines(x = ~weight, y = ~growth) |> 
  layout(yaxis = list(title_text = "Growth Rate (g/year)", 
                      range = c(0, 20),
                      showline = TRUE, showgrid = FALSE, zeroline = FALSE, mirror = TRUE),
         xaxis = list(type = "log", exponentformat = "power", 
                      title_text = "Weight (g)", 
                      showline = TRUE, showgrid = FALSE, mirror = TRUE), 
         margin = list(b = 65, l = 65))
```

### Comparative growth rate 
Comparison between the growth rate of fish in the original system to the fish in reduced resource system.

```{r message=FALSE, warning=FALSE}
gf_original <- melt(getEGrowth(params_single))
gf_original$Model <- "Original"
gf_starved <- melt(getEGrowth(params))
gf_starved$Model <- "Less prey"
gf_total <- rbind(gf_original, gf_starved)

plot_ly(gf_total) |> 
  add_lines(x = ~w, y = ~value, color = ~Model) |> 
  layout(yaxis = list(title_text = "Growth rate (g/year)",
                      showline = TRUE, showgrid = FALSE, mirror = TRUE, zeroline = FALSE,
                      range = c(0,40)),
         xaxis = list(type = "log", exponentformat = "power", 
                      title_text = "Weight (g)",
                      showline = TRUE, showgrid = FALSE, mirror = TRUE),
         margin = list(b = 65, l = 65))
```



## Biomass density spectra
### Reduced resource system
Plot the spectra to see the biomass density across weight classes of the system

```{r message=FALSE, warning=FALSE}
plotSpectra(sim_reduced, power = 2, wlim = c(1e-8, NA), ylim = c(1e-8, NA),
            time_range = 50)+
  theme_test()
```

### Comparative biomass density spectra

```{r message=FALSE, warning=FALSE}
plotSpectra2(sim_initial, name1 = "Original",
             sim_reduced, name2 = "Less prey",
             power = 2) + theme_test()
```

## Animations

We create animations to look at modelled changes over time

### Biomass abundance
```{r message=FALSE, warning=FALSE}
nf <- melt(sim_reduced@n)
n_ppf <- melt(sim_reduced@n_pp)
n_ppf$sp <- "Resource"
nf <- rbind(nf, n_ppf)

plot_ly(nf) |> 
  filter(w > 10^-5) |>  
  mutate(b = value * w^2) |> 
  add_lines(
    x = ~w, y = ~b,
    color = ~sp,
    frame = ~time,
    line = list(simplify = FALSE)) |> 
  layout(params, xaxis = list(type = "log", exponentformat = "power",
                         title_text = "body mass (g)"),
         yaxis = list(type = "log", exponentformat = "power",
                      title_text = "biomass (g/m^3)",
                      range = c(-9, 0)))
```


### Flux
```{r message=FALSE, warning=FALSE}
species <- "Target species"
time <- getTimes(sim_reduced)

flux_series <- lapply(seq_along(time), function(i) {
  n_at_t <- sim_reduced@n[i, species, ]
  data.frame(
    Time = time[i],
    Weight = w,
    Flux = n_at_t * grr)}) |>  bind_rows()

flux_series$Species <- species

plot_ly(flux_series) |> 
  add_lines(x ~Weight, 
            y ~Flux, 
            color = ~Species, 
            frame = ~Time, 
            line = list(simplify = FALSE)) |>  
  layout(params, xaxis = list(type = "log", exponentformat = "power",
                         title_text = "body mass (g)"),
         yaxis = list(type = "log", exponentformat = "power",
                      title_text = "flux (gm<sup>-2</sup>year<sup>-1</sup>)", 
                      range = c(-9, 0)))
```

## Altering fishing effort
All previous simulations have been run using maximum sustainable yield. We now alter the fishing effort to see the impact on modelled functions. 

### Flux with lower fishing effort
Effort was changed to 0.1.
```{r message=FALSE, warning=FALSE}
sim_lower <- project(params, t_max = 50, effort = 0.1)

N_lower <- finalN(sim_lower)["Target species", , drop = TRUE]
flux_lower <- grr * N_lower

lower_flux_data <- data.frame(Weight = w, 
                                Flux = flux_lower)

plot_ly(lower_flux_data) |> 
  add_lines(x = ~Weight, y = ~Flux) |> 
  layout(yaxis = list(type = "log", exponentformat = "power",
                      title_text = "Flux (gm<sup>-2</sup>year<sup>-1</sup>)", 
                      showline = TRUE, showgrid = FALSE, mirror = TRUE),
         xaxis = list(type = "log", exponentformat = "power", 
                      title_text = "Weight (g)", 
                      showline = TRUE, showgrid = FALSE, mirror = TRUE), 
         margin = list(b = 65, l = 50))
```

### Flux with higher fishing effort
Effort was changed to 1.5, which is double MSY.

```{r message=FALSE, warning=FALSE}
sim_higher <- project(params, t_max = 50, effort = 1.5)

N_higher <- finalN(sim_higher)["Target species", , drop = TRUE]
flux_higher <- grr * N_higher

higher_flux_data <- data.frame(Weight = w, 
                                Flux = flux_higher)
plot_ly(higher_flux_data) |> 
  add_lines(x = ~Weight, y = ~Flux) |> 
  layout(yaxis = list(type = "log", exponentformat = "power",
                      title_text = "Flux (gm<sup>-2</sup>year<sup>-1</sup>)", 
                      showline = TRUE, showgrid = FALSE, mirror = TRUE),
         xaxis = list(type = "log", exponentformat = "power", 
                      title_text = "Weight (g)", 
                      showline = TRUE, showgrid = FALSE, mirror = TRUE), 
         margin = list(b = 65, l = 75))
```

## Compare flux between fishing efforts

```{r message=FALSE, warning=FALSE}
reduced_flux_data$Effort = "MSY"
lower_flux_data$Effort = "Lower fishing effort"
higher_flux_data$Effort = "Higher fishing effort"


fishing_flux <- rbind(higher_flux_data, lower_flux_data, reduced_flux_data)

plot_ly(fishing_flux) |> 
  add_lines(x = ~Weight, y = ~Flux, color = ~Effort) |> 
  layout(yaxis = list(type = "log", exponentformat = "power",
                      title_text = "Flux (gm<sup>-2</sup>year<sup>-1</sup>)", 
                      showline = TRUE, showgrid = FALSE, mirror = TRUE),
         xaxis = list(type = "log", exponentformat = "power", 
                      title_text = "Weight (g)", 
                      showline = TRUE, showgrid = FALSE, mirror = TRUE), 
         margin = list(b = 65, l = 50))
```

## Yield 

### Maximum sustainable yield
We plot the yield with MSY.

```{r message=FALSE, warning=FALSE}
sim_MSY <- project(params, t_max = 150, effort = 0.75)

yield <- getYield(sim_MSY)
time_150 <- getTimes(sim_MSY)

yield_data <- data.frame(Time = time_150, 
                         Yield = yield)

plot_ly(yield_data) |> 
  filter(time_150 >= 50) |>
  add_lines(x = ~Time, y = ~Target.species) |> 
  layout(yaxis = list(type = "log", exponentformat = 'power', 
                      title_text = "Yield (g/year)",
                      showline = TRUE, showgrid = FALSE, mirror = TRUE),
         xaxis = list(title_text = "Year", 
                      showline = TRUE, showgrid = FALSE, mirror = TRUE),
         margin = list(b = 65, l = 70))
```


### Lower effort
We plot the yield with lower fishing effort.

```{r message=FALSE, warning=FALSE}
sim_low_yield <- project(params, t_max = 150, effort = 0.1)
yield_lower <- getYield(sim_low_yield)

yield_lower_data <- data.frame(Time = time_150, 
                               Yield = yield_lower)

plot_ly(yield_lower_data) |> 
  filter(time_150 >= 50) |>
  add_lines(x = ~Time, y = ~Target.species) |> 
  layout(yaxis = list(type = "log", exponentformat = 'power', 
                      title_text = "Yield (g/year)",
                      showline = TRUE, showgrid = FALSE, mirror = TRUE),
         xaxis = list(title_text = "Year", 
                      showline = TRUE, showgrid = FALSE, mirror = TRUE),
         margin = list(b = 65, l = 70))
```

### Higher fishing effort
We plot the yield with higher fishing effort.

```{r message=FALSE, warning=FALSE}
sim_high_yield <- project(params, t_max = 150, effort = 1.5)
yield_heavy <- getYield(sim_high_yield)

yield_heavy_data <- data.frame(Time = time_150, 
                         Yield = yield_heavy)

plot_ly(yield_heavy_data) |> 
  filter(time_150 >= 50) |>
  add_lines(x = ~Time, y = ~Target.species) |> 
  layout(yaxis = list(type = "log", exponentformat = 'power', 
                      title_text = "Yield (g/year)",
                      showline = TRUE, showgrid = FALSE, mirror = TRUE),
         xaxis = list(title_text = "Year", 
                      showline = TRUE, showgrid = FALSE, mirror = TRUE),
         margin = list(b = 65, l = 70))
```

### Time dependent fishing effort
We experimented with changing the fishing effort in accordance to the yield. When the peaks of yield was reached, based on the MSY graph, the effort increases to the higher fishing effort. Otherwise the effort remained at a level to produce MSY.

```{r message=FALSE, warning=FALSE}
t_total <- 500
low_effort <- 0.75
high_effort <- 1.5
threshold_value <- 0.000025
current_effort <- low_effort


sim_combined <- project(params, t_max = 1, effort = current_effort)

last_yield <- getYield(sim_combined)[idxFinalT(sim_combined), "Target species"]
if (last_yield >= threshold_value) {
  current_effort <- high_effort
} else {
  current_effort <- low_effort
}

for (t in 2:t_total) {
  sim_combined <- project(sim_combined, t_max = 1, effort = current_effort, append = TRUE)
  
  last_yield <- getYield(sim_combined)[idxFinalT(sim_combined), "Target species"]
  
  if (last_yield >= threshold_value) {
    current_effort <- high_effort
  } else {
    current_effort <- low_effort
  }
}

# Yield with changing effort
yield_fishing <- getYield(sim_combined)
time_500<- getTimes(sim_combined)

yield_fishing_data <- data.frame(Time = time_500, 
                                 Yield = yield_fishing)

plot_ly(yield_fishing_data) |> 
  filter(time_500 >= 300) |>
  add_lines(x = ~Time, y = ~Target.species) |> 
  layout(yaxis = list(type = "log", exponentformat = 'power', 
                      title_text = "Yield (g/year)",
                      showline = TRUE, showgrid = FALSE, mirror = TRUE),
         xaxis = list(title_text = "Year",
                      showline = TRUE, showgrid = FALSE, mirror = TRUE),
         margin = list(b = 65, l = 70))
```

### Changing effort

```{r message=FALSE, warning=FALSE}
effort_df <- melt(getEffort(sim_combined))

plot_ly(effort_df) |> 
  filter(time_500 >= 300) |> 
  add_lines(x = ~time, y = ~value) |> 
  layout(yaxis = list(title_text = "Effort",
                      showline = TRUE, showgrid = FALSE, mirror = TRUE),
         xaxis = list(title_text = "Time (year)", range = c(300, 500),
                      showline = TRUE, showgrid = FALSE, mirror = TRUE),
         margin = list(b = 65, l = 70))
```
