---
title: "Mizer model"
author: "Beth Sleath"
format: pdf
editor: visual
prefer-html: true
---

## Loading library

Load the packages required to design mizer model and produce figures.

```{r}
#| message: false
#| warning: false
library(mizer)
library(mizerExperimental)
library(tidyverse)
```

## Setting initial parameters

```{r}
#| message: false
#| warning: false
params_single <- newSingleSpeciesParams()

params_initial <- setResource(params_single,
                              resource_dynamics = "resource_semichemostat")

params_double <- params_initial
params_double <- setResource(params_double,
                             resource_level = 0.1)

initialN(params_double) <- 2*initialN(params_double)
initialNResource(params_double) <- initialNResource(params_double)/2

```

## Implementing fishing gear

```{r}
#| message: false
#| warning: false
gear_params(params_double) <- data.frame(
  gear = "gear",
  species = "Target species",
  catchability = 0.3,
  sel_func = "sigmoid_weight",
  sigmoidal_weight = 15,
  sigmoidal_sigma = 5)

params_double <- setFishing(params_double, gear_params = gear_params)

gear_params(params_double)
```

# Simulating ecosystem behaviour

## Flux of system

```{r}
sim_double <- project(params_double, t_max = 15, effort = 1)
```

#### Calculate flux

```{r}
N <- finalN(sim_double)["Target species", , drop = TRUE]
w <- w(params_double)

E_growth <- getEGrowth(params_double)["Target species", , drop = TRUE]
gr <- w * E_growth
flux <- gr * N
```

#### Plot flux

```{r}
#| message: false
#| warning: false
initial_flux_data <- data.frame(Weight = w, 
                             Flux = flux)
initial_flux_plot <- ggplot(initial_flux_data,
                         aes(x = Weight, y = Flux)) +
  geom_smooth() +
  scale_x_log10() +
  scale_y_log10() +
  labs(
    x = paste0("Weight (g)"),
    y = paste0("Flux (g/year)")) +
  theme_classic()
initial_flux_plot
```

## Biomass density

```{r}
plotSpectra(params_double, power = 2)
```

## Locally reduce parameters

#### Reduce resource rate 

```{r}
#| message: false
#| warning: false
rr <- resource_rate(params_double)
w_full <- w_full(params_double)
w_full[222:242]
rr[222:242] <- rr[222:242] / 100000000

rr_data <- data.frame(
  Weight = params_double@w_full,
  rr = rr)

rr_plot <- ggplot(rr_data, 
                  aes(x = Weight, y = rr)) +
  geom_smooth() +
  scale_x_log10() + 
  scale_y_log10() +
  theme_classic()
rr_plot
```

#### Reduce resource capacity

```{r}
#| message: false
#| warning: false
rc <- resource_capacity(params_double)
w_full[222:242]
rc[222:242] <- rc[222:242] / 100000000

rc_data <- data.frame(
  Weight = params_double@w_full,
  rc = rc)

rc_plot <- ggplot(rc_data, 
                  aes(x = Weight, y = rc)) +
  geom_smooth() +
  scale_x_log10() + 
  scale_y_log10() +
  theme_classic()
rc_plot
```

#### Set new parameters with local resource parameter reductions

```{r}
params_reduced <- setResource(params_double,
                              resource_capacity = rc,
                              resource_rate = rr,
                              balance = FALSE)
```

## Predation kernel 

#### Define predation kernel

```{r}
#| message: false
#| warning: false
pred_kernel <- getPredKernel(params_reduced)
pred_kernel_reduced <- pred_kernel[, 90:98, , drop = FALSE]

ggplot(melt(pred_kernel_reduced)) +
  geom_line(aes(x = w_prey, y = value)) +
  scale_x_log10(limits = c(1e-4, 100))
```

#### Narrow predation kernel

```{r}
#| message: false
#| warning: false
select(species_params(params_reduced), beta, sigma)
params <- params_reduced

given_species_params(params)$sigma <- 0.2
given_species_params(params)$beta <- 100

getPredKernel(params)[, 90:98, , drop = FALSE] %>% 
  melt() %>% 
  ggplot() +
  geom_line(aes(x = w_prey, y = value)) +
  scale_x_log10(limits = c(1e-4, 100))
```

## New flux of system

```{r}
sim_reduced <- project(params, t_max = 15, effort = 1)

N_reduced <- finalN(sim_reduced)["Target species", , drop = TRUE]
w <- w(params_reduced)

E_growth_reduced <- getEGrowth(params)["Target species", , drop = TRUE]
grr <- w * E_growth_reduced
flux_reduced <- grr * N_reduced


reduced_flux_data <- data.frame(Weight = w, 
                             Flux = flux_reduced)
```

#### Plot flux on a log-log axis

```{r}
#| message: false
#| warning: false
reduced_flux_log_plot <- ggplot(reduced_flux_data,
                         aes(x = Weight, y = Flux)) +
  geom_smooth() +
  scale_y_log10() +
  scale_x_log10() +
  labs(
    x = paste0("Weight (g)"),
    y = paste0("Flux (g/year)")) +
  theme_classic()
reduced_flux_log_plot
```

#### Plot flux on a log y-axis

```{r}
#| message: false
#| warning: false
reduced_flux_plot <- ggplot(reduced_flux_data,
                            aes(x = Weight, y = Flux)) +
  geom_smooth() +
  scale_y_log10() +
  labs(
    x = paste0("Weight (g)"),
    y = paste0("Flux (g/year)")) +
  theme_classic()
reduced_flux_plot
```

## Flux over a limited weight range

```{r}
limited_flux_data <- data.frame(Weight = w[89:101], 
                                Flux = flux_reduced[89:101])
```

#### Plot on a log-log axis

```{r}
#| message: false
#| warning: false
limited_flux_log_plot <- ggplot(limited_flux_data,
                                aes(x = Weight, y = Flux)) +
  geom_smooth() +
  scale_y_log10() +
  scale_x_log10(limits = c(25, 100)) +
  labs(
    x = paste0("Weight (g)"),
    y = paste0("Flux (g/year)")) +
  theme_classic()
limited_flux_log_plot
```

#### Plot on a log y-axis

```{r}
#| message: false
#| warning: false
limited_flux_plot <- ggplot(limited_flux_data,
                                aes(x = Weight, y = Flux)) +
  geom_smooth() +
  scale_y_log10() +
  scale_x_continuous(limits = c(25, 100)) +
  labs(
    x = paste0("Weight (g)"),
    y = paste0("Flux (g/year)")) +
  theme_classic()
limited_flux_plot
```

## Plot growth rate 

```{r}
#| message: false
#| warning: false
growth_data <- data.frame(growth = E_growth_reduced,
                          weight = w)
growth_plot <- ggplot(growth_data,
       aes(x = weight,
           y = growth)) +
  geom_smooth() +
  labs(x = paste0("Weight (g)"),
       y = paste0("Growth Rate (g/year)")) +
  theme_classic()
growth_plot
```

## Plot growth rate over a limited weight range

```{r}
#| message: false
#| warning: false
limited_growth_data <- data.frame(growth = E_growth_reduced[89:101],
                          weight = w[89:101])
limited_growth_plot <- ggplot(limited_growth_data,
                      aes(x = weight,
                          y = growth)) +
  geom_smooth() +
  labs(x = paste0("Weight (g)"),
       y = paste0("Growth Rate (g/year)")) +
  theme_classic()
limited_growth_plot
```

## Plot yield

```{r}
yield <- getYield(sim_reduced)
time <- getTimes(sim_reduced)

yield_data <- data.frame(Time = time, 
                         Yield = yield)

yield_graph <- ggplot(yield_data, 
                     aes(x = Time, y = Target.species)) + 
  geom_line() +
  labs(x = paste0("Year"),
       y = paste0("Yield (g/year)") ) +
  theme_classic()
  yield_graph
```
